---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# datamatch

<!-- badges: start -->
[![R-CMD-check](https://github.com/chross22/datamatch/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/chross22/datamatch/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of datamatch is to pull environmental data from Copernicus Marine Service and match spatially and temporally to species occurence data.

## Installation

You can install the development version of datamatch from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("chross22/datamatch")
```

## Set up

Downloads go through `copernicusmarine`, the official Python client. It is not
an R package and is not installed with this one:

``` bash
pip install copernicusmarine
copernicusmarine login
```

`login` needs a free [Copernicus Marine
account](https://data.marine.copernicus.eu/register), and only has to be run
once — the client stores the credentials itself, and this package never sees
them.

If the client is installed but R cannot find it — common when it lives in a
conda environment that RStudio does not inherit the `PATH` of — point at it
directly in `~/.Rprofile`:

``` r
options(datamatch.copernicusmarine = "~/miniconda3/bin/copernicusmarine")
```

Downloaded files are cached so a repeated request is read from disk rather than
re-fetched. The cache goes in `tools::R_user_dir("datamatch", "cache")` unless
you say otherwise:

``` r
options(datamatch.cache = "~/data/copernicus")   # or DATAMATCH_CACHE
```

## Quick start

Request variables by name and let the catalog find the product for you:

```{r quickstart, eval = FALSE}
library(datamatch)

env <- accessEnvDat(
  vars = c("SST", "SSS", "MLD"),          # no product_id, no dataset_id
  years = 2003:2017,
  months = 1:12,
  bounding_box = list(xmin = -76, xmax = -65, ymin = 35, ymax = 45)
)

matched <- matchData(speciesDat = observations, envDat = env)
```

`variable_dictionary()` lists what is available.

## Variable names

Copernicus variable codes are terse and easy to misremember — `thetao` for
temperature, `mlotst` for mixed layer depth, `zos` for sea surface height — and
getting one wrong produces a failed download rather than an obvious mistake. So
variables can be requested by name instead, and the dictionary lists what is
available, along with the product and dataset each comes from:

```{r dictionary}
library(datamatch)

variable_dictionary()
```

`as_markdown()` renders either dictionary as a pipe table, for pasting into
documentation:

```{r dictionary-md, results = "asis"}
as_markdown(variable_dictionary())
```

Pass those names to `accessEnvDat()` and the result comes back with them as
column names, rather than the Copernicus codes:

```{r named-fetch, eval = FALSE}
env <- accessEnvDat(
  vars = c("SST", "SSS", "MLD"),
  years = 2003:2017,
  months = 1:12,
  bounding_box = list(xmin = -76, xmax = -65, ymin = 35, ymax = 45)
)
names(env)
#> "SST" "SSS" "MLD" "YEAR" "MONTH" "DAY" "geometry"
```

**`product_id` and `dataset_id` can be omitted** when every variable is in the
dictionary, since the catalog already knows where they live. Raw Copernicus
codes work the same way, so `vars = c("thetao", "so")` infers its dataset too. Variables from
different datasets cannot be fetched in one request, and mixing them is refused
before any download is attempted rather than failing obscurely at the API:

```{r mixed, eval = FALSE}
accessEnvDat(vars = c("SST", "CHL"), ...)
#> Error: These variables come from different Copernicus datasets and cannot be
#>   fetched together:
#>     SST  ->  cmems_mod_glo_phy_my_0.083deg_P1M-m
#>     CHL  ->  cmems_mod_glo_bgc_my_0.25deg_P1M-m
#>   Call accessEnvDat() once per dataset.
```

Raw Copernicus codes still work exactly as before, so existing calls need no
change, and anything outside the dictionary is passed through to the API with a
warning — Copernicus serves far more than this catalog covers.

### Satellite or model?

Copernicus serves chlorophyll and primary production from two very different
sources, and both are available:

| Name | Source | Resolution | Trade-off |
|---|---|---|---|
| `CHL`, `PP` | Copernicus-GlobColour, satellite | 4 km | Observed, finer — but surface-only and gappy under persistent cloud |
| `CHL_MODEL`, `NPP_MODEL` | Biogeochemistry reanalysis | 0.25° | Gap-free and depth-resolved — but simulated, and coarser |

The plain names default to satellite, since the values are observed rather than
simulated. Switch to the model versions where cloud gaps would matter more than
resolution.

One caution: **satellite `PP` and model `NPP_MODEL` are not the same quantity.**
`PP` is depth-integrated (mg/m2/day), `NPP_MODEL` volumetric (mg/m3/day).
Substituting one for the other is a units error, not a resolution difference.

Phytoplankton functional types come from the same satellite plankton dataset as
`CHL`, so they can be fetched together:

```{r pft, eval = FALSE}
env <- accessEnvDat(
  vars = c("CHL", "DIATO", "DINO"),   # one request, one dataset
  years = 2003:2017, months = 1:12,
  bounding_box = list(xmin = -76, xmax = -65, ymin = 35, ymax = 45)
)
```

`DIATO` and `DINO` are diatom and dinophyte chlorophyll — the spring-bloom
species large copepods prefer, and the later stratified-water group
respectively.

```{r lookup, eval = FALSE}
variable_dictionary("biogeochemical")        # filter by product
variable_dataset(c("SST", "CHL"))            # which dataset each comes from
as.data.frame(variable_dictionary())$description  # full descriptions
```

## Spatial and temporal resolution

Products do not share a grid, so how resolution is handled matters:

| Source | Spatial | Temporal | Gaps |
|---|---|---|---|
| Physics reanalysis | 0.083° (~9 km) | monthly, from 1993 | gap-free |
| Biogeochemistry reanalysis | 0.25° (~28 km) | monthly, from 1993 | gap-free |
| Satellite ocean colour | 4 km | monthly, from 1997 | **cloud gaps** |
| Analysis-and-forecast | 0.083° / 0.25° | monthly, to ~10 days ahead | gap-free |

Satellite is the finest and the only observed source, but it is also the only
one with holes — and those holes are not random, clustering in the seasons and
latitudes where cloud is persistent. `fill_satellite_gaps()` substitutes the
model equivalent where the satellite saw nothing, and records the source of
every value.

**`accessEnvDat()` fetches one dataset per call**, on that dataset's native grid.
It does not resample, and it refuses to fetch variables from different datasets
together rather than quietly reconciling them.

**`matchData()` is resolution-agnostic.** Observations are matched to the nearest
environmental cell, so a 0.083-degree and a 0.25-degree product each attach
correctly to the same stations — the observation simply lands in a bigger or
smaller cell. Temporally it matches at the environmental data's own resolution,
inferred from its time steps: a monthly product matches by month, a daily one by
day. Matching a monthly mean on exact dates would match nothing, since the mean
carries one time step per month while observations fall on arbitrary days.

**Combining two products onto one grid is the caller's decision**, because
neither answer is free. Keeping the finer grid means each coarse cell's value is
repeated across the fine cells inside it: fine-scale structure survives in the
fine variables, but the coarse one is blocky rather than detailed, and a spatial
gradient computed from it measures the source grid rather than the ocean. Keeping
the coarser grid replicates nothing but discards resolution the fine variables
really had. `taupatch` exposes this as a `covariates.grid` setting.

`upscale_grid()` and `downscale_grid()` are how you act on that decision.

## Resampling

Four functions, one pair per axis. Each takes a `method`, because there is no
single right way to change resolution:

| | Coarser | Finer |
|---|---|---|
| **Space** | `upscale_grid()` | `downscale_grid()` |
| **Time** | `upscale_time()` | `downscale_time()` |

```{r resample, eval = FALSE}
chl <- accessEnvDat(vars = "CHL", years = 2010, months = 1:12, bounding_box = bb)
sst <- accessEnvDat(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)

# Satellite CHL (4 km) onto the physics grid, so the two can be modelled together
chl_on_sst <- upscale_grid(chl, to = sst)

# Or onto a stated resolution
chl_quarter <- upscale_grid(chl, to = 0.25, method = "median")

# Daily data to monthly means; monthly back out to daily
monthly <- upscale_time(daily_sst, to = "month")
daily   <- downscale_time(monthly, to = "day")
```

The target is either a resolution or **another object whose grid to adopt** —
the second form is what puts two products on one grid. A resolution is anchored
to the origin rather than to your data's corner, so two products upscaled to
0.25° land on the *same* cells instead of half a cell apart.

### Aggregating loses detail, which is the safe direction

`mean` is the default. The others exist because "the value of this coarse cell"
is not one question: `median` resists the retrieval outliers at satellite cloud
edges, `min` of depth is the shallowest point a cell contains rather than its
average, `sum` is for per-cell totals and meaningless for a concentration, and
`sd` over a period is variability as a covariate in its own right. Pass one
method for everything, or a named vector per variable:

```{r methods, eval = FALSE}
upscale_grid(env, to = 0.25, method = c(CHL = "median", DEPTH = "min"))
```

**Partly-covered periods and cells come back `NA` by default.** A mean over the
four days of January you happened to fetch is not a January mean, and nothing in
the number itself says so. `min_coverage` defaults to 0.5, measured against the
steps a period *has* — 31 for January — not the steps that were downloaded. Set
`min_coverage = 0` to aggregate whatever is present, and `keep_counts = TRUE` to
see what was behind each value.

### Interpolating adds cells, not information

This is the direction to be careful in. A 0.25° field rendered at 4 km has 4 km
*cells*, but it still resolves nothing below 0.25°, and none of these methods
consult any other data source — so there is nowhere for real fine-scale
structure to come from.

That is why the defaults are the blunt ones. `nearest` and `step` replicate the
source value, leaving the result visibly blocky at the source resolution, so the
output looks like what it is. `bilinear`, `cubic`, `linear`, and `spline` return
smooth fields that look like finely-resolved measurements and are not.

There is a further trap specific to the time axis:

> **`linear` and `spline` do not preserve the period mean.** Interpolate twelve
> monthly means to daily values, average those days back up, and you will not
> recover the months you started from — in the package's own tests a March mean
> of 15.0 comes back as 14.83. Any budget or total computed from such a series
> inherits that error. `step` preserves it exactly.

`idw` is the exception worth reaching for: alone among the spatial methods it
fills across holes rather than propagating them, which makes it useful on gappy
satellite data where `fill_satellite_gaps()` is not an option.

Resampled output keeps the `YEAR`/`MONTH`/`DAY` stamping convention, so
`matchData()` reads its resolution back correctly and can be used on the result
directly.

## Forecasts

The same variables can be requested from the analysis-and-forecast products,
which run to about ten days ahead:

```{r forecast, eval = FALSE}
env <- accessEnvDat(
  vars = c("SST", "MLD"),
  years = 2026, months = 8,
  bounding_box = list(xmin = -76, xmax = -65, ymin = 35, ymax = 45),
  mode = "forecast"
)
```

Two differences from the reanalysis are worth knowing, and are why this is a
mapping rather than a swapped identifier:

- **The forecast splits variables across more datasets.** `SST` and `SSS` share
  a dataset in the reanalysis but not in the forecast, so a set that fetches in
  one request may need several.
- **Some codes differ.** Bottom temperature is `bottomT` in the reanalysis and
  `tob` in the forecast. Requesting `BOTT` gets the right one either way.

Satellite variables have no forecast — ocean colour is observation, and there is
no observation of the future. Asking for `CHL` in forecast mode says so and
points at `CHL_MODEL`.

## Filling satellite gaps

```{r gapfill, eval = FALSE}
chl_sat <- accessEnvDat(vars = "CHL", years = 2010, months = 1:12, bounding_box = bb)
chl_mod <- accessEnvDat(vars = "CHL_MODEL", years = 2010, months = 1:12, bounding_box = bb)

filled <- fill_satellite_gaps(chl_sat, chl_mod, c(CHL = "CHL_MODEL"))
table(filled$CHL_source)
#> satellite     model   missing
#>      8214      1902         0
```

The two sources are different measurements of the same quantity, not
interchangeable ones, so the seam is left visible rather than smoothed: nothing
is rescaled by default, and a `<var>_source` column records where each value came
from. `rescale = TRUE` matches the model's mean and variance to the satellite's
over the cells where both exist, which reduces the step at the cost of altering
the model values.

## Static and basin-scale covariates

Not everything useful is a gridded Copernicus variable on a monthly time step.
Two other kinds are available, and they behave differently from the Copernicus
data and from each other.

### Seafloor terrain

Static — it does not vary by month or year — so it is fetched once for a study
area and attached to every time step. Sourced from NOAA ETOPO via `marmap`,
which is a suggested package rather than a hard dependency:

```{r bathy, eval = FALSE}
bathy <- fetch_bathymetry(
  bounding_box = list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)
)

observations <- attach_bathymetry(observations, bathy, c("DEPTH", "SLOPE", "TPI"))
```

`bathymetry_variables()` lists them: `DEPTH`, `SLOPE`, `ASPECT`, and `TPI`.

`TPI` — topographic position index — is a cell's depth relative to the mean of
the eight around it, and it answers a question depth cannot. A 100 m bank top
and a 100 m basin floor are the same depth and very different places; TPI is
positive on the first and negative on the second, and near zero on both flat
bottom and uniform slopes. It is scale-dependent by construction, describing
position within the immediate neighbourhood, so its meaning follows the
resolution of the grid it was computed on.

`attach_bathymetry()` takes either a plain data frame with coordinate columns or
an `sf` object, so it works on observations and on `accessEnvDat()` output alike.

### Climate indices

Monthly, and with **no spatial dimension** — one value describes the whole basin,
so every observation in a month receives the same number:

```{r indices, eval = FALSE}
observations <- attach_climate_index(observations, c("NAO", "AMO"))

index_dictionary()          # NAO, AO, AMO, PDO
```

That makes them a different kind of covariate from local temperature. They carry
information about *when* — what year and season it was — and none about *where*
within a region conditions are better. Useful for interannual questions ("was
this a warm-regime year?"), useless for spatial ones: a model given only indices
cannot produce a map.

One interaction to know: `attach_climate_index()` joins on year and month, and
`upscale_time(to = "year")` stamps its output `MONTH = 1`. Attaching an index to
annual data would therefore give every year January's value. Aggregate the index
to a year first instead.

## Matching to observations

`matchData()` joins environmental data to species observations at the
environmental data's own temporal resolution, inferred from its time steps. This
matters for monthly products: a monthly mean carries one time step per month
while observations fall on arbitrary days, so matching on exact dates would
match nothing.

```{r match, eval = FALSE}
matched <- matchData(speciesDat = observations, envDat = env)
```

Observations falling in a period with no environmental data are returned with
`NA` values and a warning naming the periods, rather than being dropped silently.

## Related packages

- [derivoce](https://github.com/chross22/derivoce) — derived covariates
  (gradients, FTLE/FSLE, front and isobath distances, lags, integrals) computed
  from what `accessEnvDat()` returns
