---
title: "Static and basin-scale covariates"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Static and basin-scale covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datamatch)
```

Not every useful covariate is a gridded field on a monthly step. Two other kinds
are available, and they behave differently from the gridded data and from each
other: **seafloor terrain**, which does not vary in time at all, and
**basin-scale climate indices**, which do not vary in space.

Neither is fetched with `matchData()`. They have their own attach functions,
because neither is a spatiotemporal nearest-feature join.


Not everything useful is a gridded Copernicus variable on a monthly time step.
Two other kinds are available, and they behave differently from the Copernicus
data and from each other.

### Seafloor terrain

This does not vary by month or year. So it is fetched once for a study area and
attached to every time step. It comes from NOAA ETOPO via `marmap`, a suggested
package rather than a hard dependency:

```{r bathy, eval = FALSE}
bathy <- fetch_bathymetry(
  bounding_box = list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)
)

observations <- attach_bathymetry(observations, bathy, c("DEPTH", "SLOPE", "TPI"))
```

`bathymetry_variables()` lists them: `DEPTH`, `SLOPE`, `ASPECT`, and `TPI`.

`TPI`, the topographic position index, is a cell's depth relative to the mean of
the eight around it. It answers a question depth cannot. A 100 m bank top and a
100 m basin floor are the same depth and very different places. TPI is positive
on the first, negative on the second, and near zero on both flat bottom and
uniform slopes.

It is scale-dependent by construction. It describes position within the
immediate neighbourhood, so its meaning follows the
resolution of the grid it was computed on.

`attach_bathymetry()` takes either a plain data frame with coordinate columns or
an `sf` object, so it works on observations and on `accessCopernicus()` output alike.

### Climate indices

These are monthly, and have **no spatial dimension**. One value describes the
whole basin, so every observation in a month receives the same number:

```{r indices, eval = FALSE}
observations <- attach_climate_index(observations, c("NAO", "AMO"))

index_dictionary()          # NAO, AO, AMO, PDO, LCR, AMOC
```

That makes them a different kind of covariate from local temperature. They tell
you about *when* — what year and season it was. They tell you nothing about
*where* within a region conditions are better.

So they are useful for interannual questions ("was this a warm-regime year?") and
useless for spatial ones. A model given only indices cannot produce a map.

Six are available, and they are not interchangeable.

| Name | What it measures | Units | Timescale | Record |
|---|---|---|---|---|
| `NAO` | Pressure difference, Icelandic Low to Azores High | standardized anomaly | Year to year | 1950– |
| `AO` | Strength of the polar vortex | standardized anomaly | Year to year | 1950– |
| `AMO` | Detrended North Atlantic SST anomaly | degrees C | Multidecadal | 1948– |
| `PDO` | Leading mode of North Pacific SST | standardized anomaly | Decadal | 1948– |
| `LCR` | Retroflection of the Labrador Current | fraction | Year to year | 1993–2014 |
| `AMOC` | Overturning transport at 26.5°N | Sv | Year to year | 2004– |

Most are **standardized anomalies** — in standard deviations, not anything
physical. Only `AMO` (degrees C) and `AMOC` (Sverdrups) carry real units, so a
coefficient fitted to one of those is not comparable with one fitted to `NAO`.
`index_dictionary()` carries the units at runtime.

**`NAO`** is the usual first choice in the Northwest Atlantic. It sets the
strength and track of the westerlies, and with them heat flux and mixing over the
shelves. Its winter values carry most of the signal. Summer values are noisier
and mean less.

**`AO`** is the hemispheric version of the same thing: polar vortex strength
rather than an Atlantic-specific pressure dipole. In winter it correlates
strongly with `NAO`. Including both usually buys little and costs collinearity,
so pick one unless you have a reason.

**`AMO`** is a slow background state, not a year-to-year signal. It varies over
decades. Within a study period of ten or twenty years it may act more like a
trend than a covariate. One caution: it is *derived from* North Atlantic SST, so
using it to predict local SST is partly circular.

**`PDO`** is North Pacific and included for completeness. It has limited relevance
to Atlantic shelf systems, and a relationship found with it is worth treating
sceptically.

The last two are different in kind, and get their own sections below.

#### The overturning circulation

`AMOC` is the strength of the Atlantic overturning in Sverdrups, measured
directly by the RAPID mooring array at 26.5°N. Alone among these it is a
measurement rather than a pattern derived from pressure or SST.

```{r amoc, eval = FALSE}
amoc <- fetch_climate_index("AMOC")      # monthly, April 2004 onward
observations <- attach_climate_index(observations, "AMOC")
```

Being real is also its limitation, in two ways. It **starts in April 2004**, so
it cannot reach back over a longer survey record. And it is measured **far south
of the shelf**: it describes the basin-scale circulation the Labrador and slope
currents sit within, not conditions on the Gulf of Maine or Scotian Shelf. A
weakening AMOC is associated with a warming Northwest Atlantic shelf, but that
is a chain of several links, so check the sign of any relationship against `LCR`
and `AMO` rather than assuming it.

Two practical notes. RAPID publishes this only as NetCDF, so it needs the
`ncdf4` package — a `Suggests`, installed with `install.packages("ncdf4")`. And
the published series is twelve-hourly, averaged to monthly here; the file is over
a megabyte, so it is cached like the Copernicus downloads.

> Moat BI, Smeed DA, Rayner D, Johns WE, Smith R, Volkov D, Elipot S, Petit T,
> Kajtar J, Baringer MO, Collins J (2026). Atlantic meridional overturning
> circulation observed by the RAPID-MOCHA-WBTS array at 26°N from 2004 to 2024
> (v2024.1a). British Oceanographic Data Centre, NERC, UK.
> <https://doi.org/10.5285/48d0bf43-0598-ceb2-e063-7086abc062f1>

#### Staying current

Four of the six indices are still growing. Downloads are cached, and the cache
expires on an interval matched to how often each provider actually publishes, so
a living index re-downloads on its own without being asked:

```{r indices-status, eval = FALSE}
climate_index_status()      # what is cached, how old, what is due
refresh_climate_index()     # force a re-fetch of everything still growing
refresh_climate_index("AMOC")
```

| Updates | Cache reused for | Indices |
|---|---|---|
| Monthly | 7 days | `NAO`, `AO`, `AMO`, `PDO` |
| Roughly yearly | 30 days | `AMOC` |
| Never | forever | `LCR` |

`LCR` finished at 2014, so re-downloading it cannot produce anything new and it
is skipped rather than fetched again.

Two behaviours are worth knowing, because both are deliberate.

**Staleness is judged from the data, not the cache.** A fresh download of a file
the provider stopped updating is still stale. If a series ends further behind the
present than its source's usual publishing lag, you are told, along with the
command to force a refresh and the note that if refreshing changes nothing then
the provider has not published either.

**A failed download falls back to the cached copy, with a warning.** A provider
being briefly unreachable should not become an error here when usable data is
already on disk. The warning is what keeps the old copy from being mistaken for
a current one.

#### The Labrador Current retroflection index

`LCR` is the odd one out among the five, and often the most directly useful. The
other four are atmospheric or SST patterns. This one describes a *current*: how
much of the Labrador Current turns eastward at the Grand Banks instead of
continuing southwest along the shelf.

Positive values mean stronger retroflection. That means less cold, fresh,
oxygen-rich Labrador water reaching the Scotian Shelf and Gulf of Maine. For
shelf water properties that is a shorter causal chain than the NAO, which acts on
them only indirectly through winds.

```{r lcr, eval = FALSE}
lcr <- fetch_climate_index("LCR")        # monthly, 1993-2014
observations <- attach_climate_index(observations, "LCR")
```

It is the published output of one study rather than an operational product, so
**cite it when you use it**:

> Jutras M, Dufour CO, Mucci A, Talbot LC (2023) Large-scale control of the
> retroflection of the Labrador Current. *Nature Communications* **14**:2623.
> <https://doi.org/10.1038/s41467-023-38321-y>

`index_dictionary()` prints that citation, and
`as.data.frame(index_dictionary())$reference` carries it at runtime.

Three limits worth knowing.

It **covers 1993–2014 only**, so it cannot be attached to recent observations.

The series is fetched from the paper's published source data, not recomputed. The
authors derived it by seeding 966 virtual particles per week across a line at
(53°N, 56.7°W)–(54.3°N, 52.0°W). Each was tracked for three years through
GLORYS12V1 velocities with [OceanParcels](https://parcels-code.org/). The index
is the difference between the counts crossing hydrographic sections on the
Labrador and Scotian Shelves.

Extending the record past 2014 means redoing that computation, not calling a
different function.

**The values are the raw index, not the normalized one plotted in the paper.**
The published source data runs roughly −0.09 to 0.18, consistent with a fraction
of the seeded particles. Figure 3a of the paper shows a detrended, smoothed
series normalized to [−1, 1], spanning about −0.6 to +0.5.

Both describe the same quantity, but the numbers are not comparable. Don't read a
value here against that figure. To get the paper's variant, apply its chain
yourself: detrend, 12-month rolling mean, rescale to [−1, 1], subtract the
1993–2015 mean.

One interaction to know. `attach_climate_index()` joins on year and month, and
`upscale_time(to = "year")` stamps its output `MONTH = 1`. So attaching an index
to annual data would give every year January's value. Aggregate the index to a
year first instead.
