---
title: "Getting started with datamatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with datamatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7, fig.height = 4.2)
library(datamatch)
set.seed(1)
```

This walks a set of observations from raw positions to a table a model can be
fitted to, and shows what each step actually returns.

Every number and figure below is computed when the vignette is built. Nothing is
downloaded: the covariates are synthetic, generated to look like a Gulf of Maine
grid with a seasonal cycle and satellite cloud gaps. The calls are the ones you
would make against real data, so a fetch is the only thing to substitute.

```{r fetch-for-real, eval = FALSE}
bb <- list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)

env <- accessCopernicus(vars = c("SST", "MLD"), years = 2015, months = 1:12,
                    bounding_box = bb)
```

```{r synthetic, echo = FALSE}
grid <- expand.grid(x = seq(-70, -66, by = 0.25), y = seq(41, 44, by = 0.25))

make_month <- function(m) {
  d <- grid
  # Warm in late summer, cool in late winter; colder to the north.
  d$SST <- 9 + 5.5 * sin(2 * pi * (m - 4.5) / 12) - 1.4 * (d$y - 41) +
    stats::rnorm(nrow(d), 0, 0.25)
  # Deep mixed layer in winter, shallow in summer.
  d$MLD <- 45 - 32 * sin(2 * pi * (m - 4.5) / 12) + stats::rnorm(nrow(d), 0, 3)
  d$YEAR <- 2015L
  d$MONTH <- as.integer(m)
  d$DAY <- 1L
  d
}

env <- sf::st_as_sf(do.call(rbind, lapply(1:12, make_month)),
                    coords = c("x", "y"), crs = 4326)
attr(env, "datamatch_step") <- "month"
```

`accessCopernicus()` returns one row per grid cell per time step, as an `sf` point
object:

```{r shape}
env
```

## Look before you model

Three plots answer the questions worth asking of a download, and each returns the
data it drew rather than only drawing it.

`plot_env()` maps one variable for one time step. This is the fastest way to
catch a bounding box that landed somewhere unintended, or a variable that is
entirely `NA`:

```{r plot-env}
plot_env(env, "SST", time = c(MONTH = 8))
```

`plot_series()` reduces each step to one number over the study area, so the
seasonal cycle and its spatial spread are both visible:

```{r plot-series}
series <- plot_series(env, "SST")
head(series, 4)
```

## Gaps are not spread evenly

Satellite variables are missing wherever cloud blocked the view, and those gaps
cluster in particular seasons. Adding a chlorophyll field with realistic winter
gaps:

```{r gappy}
env$CHL <- 1.2 + 2.4 * exp(-((env$MONTH - 5)^2) / 5) + stats::rnorm(nrow(env), 0, 0.1)

winter <- env$MONTH %in% c(11, 12, 1, 2)
env$CHL[winter][sample(sum(winter), round(sum(winter) * 0.8))] <- NA
env$CHL[!winter][sample(sum(!winter), round(sum(!winter) * 0.1))] <- NA
```

`plot_coverage()` is the one to run before trusting a monthly mean:

```{r plot-coverage}
coverage <- plot_coverage(env, c("SST", "CHL"))
```

```{r coverage-numbers}
round(coverage$coverage[coverage$variable == "CHL"], 2)
```

A quarter of the grid in winter and near-complete in summer is exactly the shape
that decides whether a winter value means anything.

## Matching

`matchData()` joins each observation to the nearest cell within the same time
period. Some observations here are deliberately placed in a month the covariates
do not cover, and some outside the grid:

```{r observations}
observations <- sf::st_as_sf(
  data.frame(
    lon   = c(-69.5, -68.2, -67.1, -66.5, -69.0, -64.0),
    lat   = c(41.5,  42.3,  43.1,  41.9,  43.6,  42.0),
    YEAR  = 2015L,
    MONTH = c(3L, 6L, 8L, 6L, 11L, 6L),
    count = c(12, 45, 7, 88, 3, 21)
  ),
  coords = c("lon", "lat"), crs = 4326)

matched <- matchData(observations, env)
sf::st_drop_geometry(matched)[c("MONTH", "count", "SST", "MLD", "CHL")]
```

Every observation comes back, in the same order, with the covariates attached.
The last one sits well outside the grid and still matched — `matchData()` uses
the *nearest* cell, and nearest has no maximum distance. That is worth knowing:
an observation far outside the study area is joined to the closest edge cell
rather than dropped, so check your bounding box covers your stations.

The November observation has `NA` chlorophyll, because that cell was under cloud.
That is a real gap rather than a failure, and the source of it is worth checking
before dropping the row:

```{r missing}
colSums(is.na(sf::st_drop_geometry(matched)))
```

## Choosing a resolution

Real products do not share a grid. Physics is 0.083°, biogeochemistry 0.25°, and
satellite ocean colour 4 km, so combining them means deciding which grid to keep.

Aggregating is the safe direction, because every value in the result summarises
values that were really measured:

```{r upscale}
coarse <- upscale_grid(env, to = 0.5, vars = "SST", min_coverage = 0)

c(cells_before = nrow(env) / 12, cells_after = nrow(coarse) / 12)
```

Interpolating is the direction to be careful in. It adds cells, not information:

```{r downscale}
fine <- downscale_grid(coarse, to = 0.25, vars = "SST")

# `nearest`, the default, invents no values: every one was already in the source.
all(stats::na.omit(fine$SST) %in% coarse$SST)
```

`bilinear` would return a smooth field that looks like a finely-resolved
measurement and is not, which is why the blunt method is the default.

### Partial coverage comes back NA

Aggregating chlorophyll over the winter, when most cells are empty, is the case
`min_coverage` exists for:

```{r min-coverage}
january <- env[env$MONTH == 1, ]

strict <- upscale_grid(january, to = 0.5, vars = "CHL", min_coverage = 0.5)
loose  <- upscale_grid(january, to = 0.5, vars = "CHL", min_coverage = 0)

c(reported_at_half_coverage = sum(!is.na(strict$CHL)),
  reported_with_no_guard    = sum(!is.na(loose$CHL)))
```

Both numbers are available; only one of them says how much went into it.
`keep_counts = TRUE` returns the fraction behind each value.

## Covariates that are not gridded

Two other kinds attach differently. Seafloor terrain is static, so the same
value goes to every time step at a location:

```{r bathymetry, eval = FALSE}
bathy <- fetch_bathymetry(bounding_box = bb)
matched <- attach_bathymetry(matched, bathy, c("DEPTH", "SLOPE", "TPI"))
```

Climate indices have no spatial dimension at all — one value per month describes
the whole basin, so every observation in a month receives the same number:

```{r indices, eval = FALSE}
matched <- attach_climate_index(matched, c("NAO", "AMOC"))
```

That makes them a different kind of covariate. They carry information about
*when*, and none about *where* within a region conditions are better. A model
given only indices cannot produce a map.

```{r index-table}
as.data.frame(index_dictionary())[c("name", "units", "source")]
```

## Where to go next

- `vignette("datamatch")` is this document; the
  [README](https://github.com/chross22/datamatch) covers the same ground in more
  depth, including forecasts, daily data, and gap filling.
- `variable_dictionary()` and `index_dictionary()` list what can be fetched.
- The README's References section lists the DOI for every data source, since the
  obligation to cite travels with the data.
