---
title: "Choosing a data source"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a data source}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datamatch)
```

datamatch fetches from five sources and joins any of them to your observations
the same way. This is about which one to reach for, and what changes when you
do.

Nothing here is downloaded when the vignette is built — the fetches are shown
but not run, because each needs a network and some need several gigabytes.

## The five, side by side

| | `accessCopernicus()` | `accessFVCOM()` | `accessHYCOM()` | `accessCCMP()` | `accessERDDAP()` |
|---|---|---|---|---|---|
| Source | Copernicus Marine | NECOFS / any FVCOM | HYCOM GOFS 3.1 | RSS CCMP v3.1 | NOAA ERDDAP |
| Kind | global reanalysis, forecast, satellite | regional coastal model | global model | wind analysis | satellite analysis |
| Grid | 0.083°–4 km, regular | unstructured mesh | 0.08° regular | 0.25° regular | 0.01°–0.04° regular |
| Steps | monthly, daily, hourly | monthly (GOM3), hourly (GOM7) | 3-hourly | 6-hourly | daily |
| Record | 1993– | 1978–2013, then 2025– | 1994–2024 across archives | 1993–present | 2002– (SST), 2012– (chl) |
| Bottom salinity | derived | free | free | — | — |
| Wind stress | yes | GOM3 only | — | — | — |
| Subsetting | server-side | client-side | server-side | **whole globe** | server-side |
| Account needed | yes | no | no | no | no |

Two of those records are chains rather than one run, and the gaps matter more
than the endpoints. **FVCOM has nothing between 2014 and 2024** — the GOM3
hindcast stops in 2013 and the GOM7 forecast archive starts in 2025, on a
different mesh. **HYCOM reaches 2024 only by crossing from the reanalysis into
a series of operational experiments**, which is a seam in how the values were
produced. `hycom_covering(date)` says which archives hold a given day.

```{r catalogs, eval = FALSE}
variable_dictionary()   # Copernicus
fvcom_dictionary()      # FVCOM
hycom_dictionary()      # HYCOM
ccmp_dictionary()       # CCMP
erddap_dictionary()     # MUR and VIIRS
```

## They share variable names on purpose

A covariate arrives in a column of the same name whichever source supplied it,
so everything downstream — `matchData()`, resampling, plotting, and whatever
model you fit — works unchanged:

```{r shared, eval = FALSE}
bb <- list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)

copernicus <- accessCopernicus(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
fvcom      <- accessFVCOM(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
hycom      <- accessHYCOM(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
```

All three produce an `SST` column. **They are three different models and three
different numbers.** Sharing the name makes them mechanically interchangeable,
which is the point; it does not make them scientifically interchangeable, which
is why it is worth recording which you used.

That property is useful deliberately. Fetching the same variable from two
sources and comparing them is a real check on a result:

```{r compare, eval = FALSE}
matched <- matchData(observations, copernicus)          # SST
matched <- matchData(matched, hycom)                    # SST.matched

plot(matched$SST, matched$SST.matched)
```

A colliding name is suffixed `.matched` rather than overwriting yours, so both
survive the join and the disagreement is visible.

## Which to reach for

**Start with Copernicus.** It has the widest variable list — physics,
biogeochemistry, satellite ocean colour, wind and stress — the longest record
with a forecast, and server-side subsetting. Everything else here is for
something Copernicus does not do well.

**Reach for FVCOM when the coast is the point.** A Gulf of Maine box holding
1,742 GLORYS cells holds 6,579 GOM3 nodes, concentrated where the bathymetry is
complicated. If your question is about a shelf, a bank, or a channel rather than
a basin, that resolution is the reason. The hindcast ends in 2013; the GOM7
forecast archive picks up in 2025 on a different mesh, with nothing in between.

**Reach for HYCOM for bottom fields, or for a second opinion.** It publishes
`salinity_bottom` and `water_temp_bottom` as real fields where Copernicus has no
bottom salinity at all, and being an independent model it makes agreement
meaningful. The reanalysis covers 1994–2015; operational archives carry it to
September 2024.

**Reach for CCMP for winds over a long record.** Six-hourly from 1993 to within
days of the present, where the Copernicus wind is monthly from mid-1994 or
hourly only from 2007. But it carries no stress.

**Reach for ERDDAP for resolution.** MUR is 0.01° — about a kilometre, against
nine for the physics reanalysis — and needs no account, where the same product
at PO.DAAC needs an Earthdata login. It is a satellite analysis of the
*foundation* temperature rather than a model level, which is a different
quantity from a model `SST`, not a better measurement of the same one.

## Bottom salinity, four ways

`BOTS` is the clearest case of the same name costing different amounts:

```{r bots, eval = FALSE}
# Copernicus reanalysis: derived. Fetches ~50 depth levels to keep the deepest
# wet one, so it must be fetched alone, and reports the depth it used.
bots <- accessCopernicus(vars = "BOTS", years = 2010, months = 1:12, bounding_box = bb)
bots$BOTS_depth

# Copernicus forecast: published outright as `sob`, no derivation.
accessCopernicus(vars = c("BOTT", "BOTS"), years = 2026, months = 7,
             bounding_box = bb, mode = "forecast")

# FVCOM: the deepest sigma layer is the sea floor everywhere. Free.
accessFVCOM(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)

# HYCOM: `salinity_bottom` is its own field. Free.
accessHYCOM(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)
```

Only the first is expensive, and only the first returns `BOTS_depth` — because
only the first had to choose a level. The deepest wet model level is not the sea
floor, and in deep water sits well above it, so the depth is reported rather
than left to assume.

## Sub-daily data, and means that are not means

Four of the five publish below daily somewhere, and none of them publishes a
daily mean:

| Source | Native step | `frequency` |
|---|---|---|
| Copernicus wind | hourly | `"hourly"` |
| FVCOM `GOM7` | hourly | `"hourly"`, or `"daily"` for a snapshot |
| HYCOM | 3-hourly | `"3hourly"`, or `"daily"` for a snapshot |
| CCMP | 6-hourly | `"6hourly"`, or `"daily"` for a snapshot |

Where a source has no mean, none is invented. `frequency = "daily"` on HYCOM and
CCMP takes one **snapshot** at a chosen hour — an instant, not an average. A
real mean is `upscale_time()`'s job, which keeps the aggregation visible and the
choice of summary yours:

```{r subdaily, eval = FALSE}
steps <- accessCCMP(vars = c("UWND", "VWND"), frequency = "6hourly",
                    dates = "2010-06-15", bounding_box = bb)

mean_wind <- upscale_time(steps, to = "day")
peak_wind <- upscale_time(steps, to = "day", method = "max")
```

The coverage check knows each source's own step, so a complete CCMP day scores
4/4 rather than 4/24.

One caution that is easy to miss: **a mean of the components is not a mean
speed.** Averaging `UWND` and `VWND` over a day and taking the magnitude gives
the net displacement of air; averaging `WSPD` gives how hard it blew. On a day
the wind reversed, the first is near zero and the second is not.

## Longitude, and one trap handled for you

Pass `bounding_box` with longitudes **negative west** to every function here.
CCMP is stored on a 0–360 grid and is converted on the way in and back on the
way out, so its results overlay the others without adjustment. You should never
have to think about it — but if you fetch CCMP by hand elsewhere, that is the
difference between the Gulf of Maine and the Indian Ocean.

## Putting several together

The pattern is the same as chaining two Copernicus products: each call adds
columns and leaves the row count alone.

```{r together, eval = FALSE}
matched <- matchData(observations, accessCopernicus(vars = c("SST", "MLD"),
                                                years = 2010, months = 1:12,
                                                bounding_box = bb))
matched <- matchData(matched, accessHYCOM(vars = "BOTS", years = 2010,
                                          months = 1:12, bounding_box = bb))
matched <- matchData(matched, accessCCMP(vars = "WSPD", years = 2010,
                                         months = 1:12, bounding_box = bb))

matched <- attach_bathymetry(matched, fetch_bathymetry(bounding_box = bb),
                             c("DEPTH", "SLOPE"))
matched <- attach_climate_index(matched, "NAO")

colSums(is.na(sf::st_drop_geometry(matched)))
```

Each source fails differently at the edges of its record — FVCOM after 2013,
HYCOM after 2015, `LCR` after 2014 — so that last line is worth running before
fitting anything.

## Citing what you used

Every source here is somebody else's work, and the obligation travels with the
data rather than with this package:

```{r cite, eval = FALSE}
fvcom_archives()$GOM3$reference
hycom_archives()$GLBv53X$reference
ccmp_versions()$`v03.1`$reference
index_dictionary()          # carries the climate index references
```

The README's References section lists the Copernicus product DOIs. Cite whichever
you actually fetched from.

## Satellite or model?

Copernicus serves chlorophyll and primary production from two very different
sources, and both are available:

| Name | Source | Resolution | Trade-off |
|---|---|---|---|
| `CHL`, `PP` | Copernicus-GlobColour, satellite | 4 km | Observed, finer — but surface-only and gappy under persistent cloud |
| `CHL_MODEL`, `NPP_MODEL` | Biogeochemistry reanalysis | 0.25° | Gap-free and depth-resolved — but simulated, and coarser |

The plain names default to satellite, since the values are observed rather than
simulated. Switch to the model versions where cloud gaps would matter more than
resolution.

One caution: **satellite `PP` and model `NPP_MODEL` are not the same quantity.**
`PP` is depth-integrated (mg/m2/day), `NPP_MODEL` volumetric (mg/m3/day).
Substituting one for the other is a units error, not a resolution difference.

Phytoplankton functional types come from the same satellite plankton dataset as
`CHL`, so they can be fetched together:

```{r pft, eval = FALSE}
env <- accessCopernicus(
  vars = c("CHL", "DIATO", "DINO"),   # one request, one dataset
  years = 2003:2017, months = 1:12,
  bounding_box = list(xmin = -76, xmax = -65, ymin = 35, ymax = 45)
)
```

`DIATO` and `DINO` are diatom and dinophyte chlorophyll — the spring-bloom
species large copepods prefer, and the later stratified-water group
respectively.

## Bottom salinity

`BOTS` pairs with `BOTT`, but it is not fetched the same way, because
**GLORYS12V1 does not publish it.** The reanalysis has temperature at the sea
floor and no salinity counterpart. So it is derived: the full salinity column is
fetched and the deepest wet level in each cell kept.

```{r bots-readme, eval = FALSE}
bots <- accessCopernicus(vars = "BOTS", years = 2010:2014, months = 1:12,
                     bounding_box = bb)
#> BOTS is not published by this product. Deriving it from the full 'so' column
#> and keeping the deepest wet level in each cell; the depth used comes back as
#> BOTS_depth.
```

The depth each value came from is returned as `BOTS_depth` rather than left to
be assumed. That matters because the deepest wet *model level* is not the sea
floor: level spacing coarsens with depth, so in deep water the value can sit a
long way above the bottom. In shelf water it is within a few metres.

Two consequences, both deliberate:

- **It must be fetched on its own.** The whole depth column is a different
  request from the single level `SST` wants, so mixing them is an error rather
  than a quiet second download. Call twice and chain `matchData()`.
- **It costs far more.** Roughly fifty levels are downloaded over the same box to
  keep one, so a large box is much slower than the same box of `SST`.

In forecast mode none of this applies. The analysis-and-forecast product
publishes sea-floor salinity outright, so `BOTS` is an ordinary variable there,
fetches alongside `BOTT`, and returns no `BOTS_depth`:

```{r bots-forecast, eval = FALSE}
accessCopernicus(vars = c("BOTT", "BOTS"), years = 2026, months = 8,
             bounding_box = bb, mode = "forecast")
```

The two are the same quantity by construction but not the same number — one is
the deepest level of a 50-level grid, the other Copernicus's own diagnostic — so
a record spanning both modes has a seam in it.

```{r lookup, eval = FALSE}
variable_dictionary("biogeochemical")        # filter by product
variable_dictionary("wind")                  # or just the winds
fvcom_dictionary()                           # the FVCOM catalog, for accessFVCOM()
fvcom_archives()                             # which FVCOM archives are built in
hycom_dictionary()                           # the HYCOM catalog, for accessHYCOM()
hycom_archives()                             # which HYCOM archives can be read
hycom_covering("2019-06-15")                 # which of them span a given date
ccmp_dictionary()                            # the CCMP catalog, for accessCCMP()
ccmp_versions()                              # which CCMP versions can be read
erddap_dictionary()                          # MUR and VIIRS, for accessERDDAP()
erddap_datasets()                            # which ERDDAP datasets ship
variable_dataset(c("SST", "CHL"))            # which dataset each comes from
as.data.frame(variable_dictionary())$description  # full descriptions
```
