---
title: "Choosing a data source"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a data source}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datamatch)
```

datamatch fetches from four sources and joins any of them to your observations
the same way. This is about which one to reach for, and what changes when you
do.

Nothing here is downloaded when the vignette is built — the fetches are shown
but not run, because each needs a network and some need several gigabytes.

## The four, side by side

| | `accessEnvDat()` | `accessFVCOM()` | `accessHYCOM()` | `accessCCMP()` |
|---|---|---|---|---|
| Source | Copernicus Marine | NECOFS / any FVCOM | HYCOM GOFS 3.1 | RSS CCMP v3.1 |
| Kind | global reanalysis, forecast, satellite | regional coastal model | global model | wind analysis |
| Grid | 0.083°–4 km, regular | unstructured mesh | 0.08° regular | 0.25° regular |
| Steps | monthly, daily, hourly | monthly | 3-hourly | 6-hourly |
| Record | 1993– | 1978–2013 | 1994–2015 | 1993–present |
| Bottom salinity | derived | free | free | — |
| Wind stress | yes | yes (model forcing) | — | — |
| Subsetting | server-side | client-side | server-side | **whole globe** |

```{r catalogs, eval = FALSE}
variable_dictionary()   # Copernicus
fvcom_dictionary()      # FVCOM
hycom_dictionary()      # HYCOM
ccmp_dictionary()       # CCMP
```

## They share variable names on purpose

A covariate arrives in a column of the same name whichever source supplied it,
so everything downstream — `matchData()`, resampling, plotting, and whatever
model you fit — works unchanged:

```{r shared, eval = FALSE}
bb <- list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)

copernicus <- accessEnvDat(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
fvcom      <- accessFVCOM(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
hycom      <- accessHYCOM(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
```

All three produce an `SST` column. **They are three different models and three
different numbers.** Sharing the name makes them mechanically interchangeable,
which is the point; it does not make them scientifically interchangeable, which
is why it is worth recording which you used.

That property is useful deliberately. Fetching the same variable from two
sources and comparing them is a real check on a result:

```{r compare, eval = FALSE}
matched <- matchData(observations, copernicus)          # SST
matched <- matchData(matched, hycom)                    # SST.matched

plot(matched$SST, matched$SST.matched)
```

A colliding name is suffixed `.matched` rather than overwriting yours, so both
survive the join and the disagreement is visible.

## Which to reach for

**Start with Copernicus.** It has the widest variable list — physics,
biogeochemistry, satellite ocean colour, wind and stress — the longest record
with a forecast, and server-side subsetting. Everything else here is for
something Copernicus does not do well.

**Reach for FVCOM when the coast is the point.** A Gulf of Maine box holding
about 2,700 GLORYS cells holds some 19,000 GOM3 nodes, concentrated where the
bathymetry is complicated. If your question is about a shelf, a bank, or a
channel rather than a basin, that resolution is the reason. It ends in 2013.

**Reach for HYCOM for bottom fields, or for a second opinion.** It publishes
`salinity_bottom` and `water_temp_bottom` as real fields where Copernicus has no
bottom salinity at all, and being an independent model it makes agreement
meaningful.

**Reach for CCMP for winds over a long record.** Six-hourly from 1993 to within
days of the present, where the Copernicus wind is monthly from mid-1994 or
hourly only from 2007. But it carries no stress.

## Bottom salinity, four ways

`BOTS` is the clearest case of the same name costing different amounts:

```{r bots, eval = FALSE}
# Copernicus reanalysis: derived. Fetches ~50 depth levels to keep the deepest
# wet one, so it must be fetched alone, and reports the depth it used.
bots <- accessEnvDat(vars = "BOTS", years = 2010, months = 1:12, bounding_box = bb)
bots$BOTS_depth

# Copernicus forecast: published outright as `sob`, no derivation.
accessEnvDat(vars = c("BOTT", "BOTS"), years = 2026, months = 7,
             bounding_box = bb, mode = "forecast")

# FVCOM: the deepest sigma layer is the sea floor everywhere. Free.
accessFVCOM(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)

# HYCOM: `salinity_bottom` is its own field. Free.
accessHYCOM(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)
```

Only the first is expensive, and only the first returns `BOTS_depth` — because
only the first had to choose a level. The deepest wet model level is not the sea
floor, and in deep water sits well above it, so the depth is reported rather
than left to assume.

## Sub-daily data, and means that are not means

Three of the four publish below daily, and none of them publishes a daily mean:

| Source | Native step | `frequency` |
|---|---|---|
| Copernicus wind | hourly | `"hourly"` |
| HYCOM | 3-hourly | `"3hourly"`, or `"daily"` for a snapshot |
| CCMP | 6-hourly | `"6hourly"`, or `"daily"` for a snapshot |

Where a source has no mean, none is invented. `frequency = "daily"` on HYCOM and
CCMP takes one **snapshot** at a chosen hour — an instant, not an average. A
real mean is `upscale_time()`'s job, which keeps the aggregation visible and the
choice of summary yours:

```{r subdaily, eval = FALSE}
steps <- accessCCMP(vars = c("UWND", "VWND"), frequency = "6hourly",
                    dates = "2010-06-15", bounding_box = bb)

mean_wind <- upscale_time(steps, to = "day")
peak_wind <- upscale_time(steps, to = "day", method = "max")
```

The coverage check knows each source's own step, so a complete CCMP day scores
4/4 rather than 4/24.

One caution that is easy to miss: **a mean of the components is not a mean
speed.** Averaging `UWND` and `VWND` over a day and taking the magnitude gives
the net displacement of air; averaging `WSPD` gives how hard it blew. On a day
the wind reversed, the first is near zero and the second is not.

## Longitude, and one trap handled for you

Pass `bounding_box` with longitudes **negative west** to every function here.
CCMP is stored on a 0–360 grid and is converted on the way in and back on the
way out, so its results overlay the others without adjustment. You should never
have to think about it — but if you fetch CCMP by hand elsewhere, that is the
difference between the Gulf of Maine and the Indian Ocean.

## Putting several together

The pattern is the same as chaining two Copernicus products: each call adds
columns and leaves the row count alone.

```{r together, eval = FALSE}
matched <- matchData(observations, accessEnvDat(vars = c("SST", "MLD"),
                                                years = 2010, months = 1:12,
                                                bounding_box = bb))
matched <- matchData(matched, accessHYCOM(vars = "BOTS", years = 2010,
                                          months = 1:12, bounding_box = bb))
matched <- matchData(matched, accessCCMP(vars = "WSPD", years = 2010,
                                         months = 1:12, bounding_box = bb))

matched <- attach_bathymetry(matched, fetch_bathymetry(bounding_box = bb),
                             c("DEPTH", "SLOPE"))
matched <- attach_climate_index(matched, "NAO")

colSums(is.na(sf::st_drop_geometry(matched)))
```

Each source fails differently at the edges of its record — FVCOM after 2013,
HYCOM after 2015, `LCR` after 2014 — so that last line is worth running before
fitting anything.

## Citing what you used

Every source here is somebody else's work, and the obligation travels with the
data rather than with this package:

```{r cite, eval = FALSE}
fvcom_archives()$GOM3$reference
hycom_archives()$GLBv53X$reference
ccmp_versions()$`v03.1`$reference
index_dictionary()          # carries the climate index references
```

The README's References section lists the Copernicus product DOIs. Cite whichever
you actually fetched from.
