---
title: "Choosing a data source"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing a data source}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datamatch)
```

datamatch fetches from seven sources and joins any of them to your observations
the same way. This is about which one to reach for, and what changes when you
do.

Nothing here is downloaded when the vignette is built — the fetches are shown
but not run, because each needs a network and some need several gigabytes.

## The seven, side by side

The four models first:

| | `accessCopernicus()` | `accessFVCOM()` | `accessHYCOM()` | `accessCEFI()` |
|---|---|---|---|---|
| Source | Copernicus Marine | NECOFS / any FVCOM | HYCOM GOFS 3.1 | NOAA CEFI, MOM6-COBALT |
| Kind | global reanalysis, forecast, satellite | regional coastal model | global model | regional coupled model |
| Grid | 0.083°–4 km, regular | unstructured mesh | 0.08° regular | 0.083° regular (regridded) |
| Extent | global | Gulf of Maine | global | NW Atlantic only |
| Steps | monthly, daily, hourly | monthly (GOM3), hourly (GOM7) | 3-hourly | monthly, daily |
| Record | 1993– | 1978–2013, then 2025– | 1994–2024 across archives | 1993–2023 |
| Biogeochemistry | 0.25°, separate product | — | — | **same grid as physics** |
| Bottom salinity | derived | free | free | free |
| Wind stress | yes | GOM3 only | — | — |
| Subsetting | server-side | client-side | server-side | server-side |
| Account needed | yes | no | no | no |

Then the three observational products:

| | `accessCCMP()` | `accessERDDAP()` | `accessOBDAAC()` |
|---|---|---|---|
| Source | RSS CCMP v3.1 | NOAA ERDDAP | NASA OB.DAAC |
| Kind | wind analysis | satellite analysis | satellite Level-3 composites |
| Grid | 0.25° regular | 0.01°–0.04° regular | 4 km or 9 km regular |
| Steps | 6-hourly | daily | daily, monthly |
| Record | 1993–present | 2002– (SST), 2012– (chl) | **1997–** |
| Gaps under cloud | n/a | none (analysed) | **yes, single-sensor** |
| Subsetting | **whole globe** | server-side | **whole globe** |
| Account needed | no | no | **yes, Earthdata** |

Two of those records are chains rather than one run, and the gaps matter more
than the endpoints. **FVCOM has nothing between 2014 and 2024** — the GOM3
hindcast stops in 2013 and the GOM7 forecast archive starts in 2025, on a
different mesh. **HYCOM reaches 2024 only by crossing from the reanalysis into
a series of operational experiments**, which is a seam in how the values were
produced. `hycom_covering(date)` says which archives hold a given day, and
`accessHYCOM(archive = "continuous")` reads across them in one call — warning
once at the crossing, and recording on every row which run it came from, since
a step in a series across that date can be the change of run rather than the
ocean.

```{r catalogs, eval = FALSE}
variable_dictionary()   # Copernicus
fvcom_dictionary()      # FVCOM
hycom_dictionary()      # HYCOM
cefi_dictionary()       # CEFI
ccmp_dictionary()       # CCMP
erddap_dictionary()     # MUR and VIIRS
obdaac_dictionary()     # SeaWiFS, MODIS, VIIRS
```

CEFI and OB.DAAC are the two most recently added, and each is here for one
thing the other six do not do. CEFI is the only source carrying **physics and
biogeochemistry on one grid at one resolution**, which is what a question about
nutrients or oxygen alongside temperature actually needs. OB.DAAC is the only
one reaching back to **1997**, which is what a question about trend rather than
state needs.

## They share variable names on purpose

A covariate arrives in a column of the same name whichever source supplied it,
so everything downstream — `matchData()`, resampling, plotting, and whatever
model you fit — works unchanged:

```{r shared, eval = FALSE}
bb <- list(xmin = -70, xmax = -66, ymin = 41, ymax = 44)

copernicus <- accessCopernicus(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
fvcom      <- accessFVCOM(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
hycom      <- accessHYCOM(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)
```

All three produce an `SST` column. **They are three different models and three
different numbers.** Sharing the name makes them mechanically interchangeable,
which is the point; it does not make them scientifically interchangeable, which
is why it is worth recording which you used.

That property is useful deliberately. Fetching the same variable from two
sources and comparing them is a real check on a result:

```{r compare, eval = FALSE}
matched <- matchData(observations, copernicus)          # SST
matched <- matchData(matched, hycom)                    # SST.matched

plot(matched$SST, matched$SST.matched)
```

A colliding name is suffixed `.matched` rather than overwriting yours, so both
survive the join and the disagreement is visible.

## Which to reach for

**Start with Copernicus.** It has the widest variable list — physics,
biogeochemistry, satellite ocean colour, wind and stress — the longest record
with a forecast, and server-side subsetting. Everything else here is for
something Copernicus does not do well.

**Reach for FVCOM when the coast is the point.** A Gulf of Maine box holding
1,742 GLORYS cells holds 6,579 GOM3 nodes, concentrated where the bathymetry is
complicated. If your question is about a shelf, a bank, or a channel rather than
a basin, that resolution is the reason. The hindcast ends in 2013; the GOM7
forecast archive picks up in 2025 on a different mesh, with nothing in between.

**Reach for HYCOM for bottom fields, or for a second opinion.** It publishes
`salinity_bottom` and `water_temp_bottom` as real fields where Copernicus has no
bottom salinity at all, and being an independent model it makes agreement
meaningful. The reanalysis covers 1994–2015; operational archives carry it to
September 2024.

**Reach for CCMP for winds over a long record.** Six-hourly from 1993 to within
days of the present, where the Copernicus wind is monthly from mid-1994 or
hourly only from 2007. But it carries no stress.

**Reach for ERDDAP for resolution.** MUR is 0.01° — about a kilometre, against
nine for the physics reanalysis — and needs no account, where the same product
at PO.DAAC needs an Earthdata login. It is a satellite analysis of the
*foundation* temperature rather than a model level, which is a different
quantity from a model `SST`, not a better measurement of the same one.

**Reach for CEFI for biogeochemistry on the shelf.** It is the only source here
where nitrate, oxygen, pH, pCO2, phytoplankton carbon and mesozooplankton
biomass sit on the same twelfth-degree grid as the temperature and salinity —
the Copernicus biogeochemical reanalysis is a quarter degree and a separate
product, so pairing them means regridding across a factor of three. CEFI also
publishes sea-floor salinity outright rather than by derivation. Its cost is
extent: the Northwest Atlantic and nothing else, so a box outside roughly
98°W–36°W and 5°N–58°N is refused. Its daily output is biogeochemistry only.

**Reach for OB.DAAC for the length of the ocean colour record.** SeaWiFS begins
in September 1997, where the ERDDAP chlorophyll entries begin in 2012 — the
difference between a decade of chlorophyll and nearly three. It carries what
ocean colour measures besides chlorophyll (`KD490`, `PAR`, `POC`, `PIC`) from
the same retrieval, and both MODIS instruments alongside all three VIIRS so
there is a sensor either side of any gap you need to bridge. Its costs are an
Earthdata Login, whole-globe downloads, and cloud: these are single-sensor
composites, so a daily field outside the tropics is mostly `NA`. Composite to
monthly, or use `fill_satellite_gaps()`.

Two warnings that belong with it. The **sensors are not interchangeable** —
where they overlap they disagree, and a series stitched across a mission
boundary has an instrumental step in it; where a long consistent record matters
more than any one sensor, the Copernicus-GlobColour entries are multi-sensor and
built for that. And **PACE is not among the sensors**, deliberately: its
standard Level-3 mapped catalog publishes no chlorophyll suite, and its
filenames carry a processing version that changes with each reprocessing and
cannot be constructed. `?obdaac_sensors` gives the detail.

## Bottom salinity, five ways

`BOTS` is the clearest case of the same name costing different amounts:

```{r bots, eval = FALSE}
# Copernicus reanalysis: derived. Fetches ~50 depth levels to keep the deepest
# wet one, so it must be fetched alone, and reports the depth it used.
bots <- accessCopernicus(vars = "BOTS", years = 2010, months = 1:12, bounding_box = bb)
bots$BOTS_depth

# Copernicus forecast: published outright as `sob`, no derivation.
accessCopernicus(vars = c("BOTT", "BOTS"), years = 2026, months = 7,
             bounding_box = bb, mode = "forecast")

# FVCOM: the deepest sigma layer is the sea floor everywhere. Free.
accessFVCOM(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)

# HYCOM: `salinity_bottom` is its own field. Free.
accessHYCOM(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)

# CEFI: `sob` is the model's own sea-floor diagnostic. Free, and NW Atlantic only.
accessCEFI(vars = c("BOTT", "BOTS"), years = 2010, months = 1:12, bounding_box = bb)
```

Only the first is expensive, and only the first returns `BOTS_depth` — because
only the first had to choose a level. The deepest wet model level is not the sea
floor, and in deep water sits well above it, so the depth is reported rather
than left to assume.

## Sub-daily data, and means that are not means

Four of the seven publish below daily somewhere, and none of them publishes a
daily mean:

| Source | Native step | `frequency` |
|---|---|---|
| Copernicus wind | hourly | `"hourly"` |
| FVCOM `GOM7` | hourly | `"hourly"`, or `"daily"` for a snapshot |
| HYCOM | 3-hourly | `"3hourly"`, or `"daily"` for a snapshot |
| CCMP | 6-hourly | `"6hourly"`, or `"daily"` for a snapshot |

The other three stop at daily. CEFI is monthly or daily and nothing finer;
ERDDAP is daily; OB.DAAC is daily or a monthly composite, and its eight-day
composite is refused because `matchData()` has no step that means eight days.

Where a source has no mean, none is invented. `frequency = "daily"` on HYCOM and
CCMP takes one **snapshot** at a chosen hour — an instant, not an average. A
real mean is `upscale_time()`'s job, which keeps the aggregation visible and the
choice of summary yours:

```{r subdaily, eval = FALSE}
steps <- accessCCMP(vars = c("UWND", "VWND"), frequency = "6hourly",
                    dates = "2010-06-15", bounding_box = bb)

mean_wind <- upscale_time(steps, to = "day")
peak_wind <- upscale_time(steps, to = "day", method = "max")
```

The coverage check knows each source's own step, so a complete CCMP day scores
4/4 rather than 4/24.

One caution that is easy to miss: **a mean of the components is not a mean
speed.** Averaging `UWND` and `VWND` over a day and taking the magnitude gives
the net displacement of air; averaging `WSPD` gives how hard it blew. On a day
the wind reversed, the first is near zero and the second is not.

## Longitude, and one trap handled for you

Pass `bounding_box` with longitudes **negative west** to every function here.
CCMP is stored on a 0–360 grid and is converted on the way in and back on the
way out, so its results overlay the others without adjustment. You should never
have to think about it — but if you fetch CCMP by hand elsewhere, that is the
difference between the Gulf of Maine and the Indian Ocean.

## Putting several together

The pattern is the same as chaining two Copernicus products: each call adds
columns and leaves the row count alone.

```{r together, eval = FALSE}
matched <- matchData(observations, accessCopernicus(vars = c("SST", "MLD"),
                                                years = 2010, months = 1:12,
                                                bounding_box = bb))
matched <- matchData(matched, accessHYCOM(vars = "BOTS", years = 2010,
                                          months = 1:12, bounding_box = bb))
matched <- matchData(matched, accessCCMP(vars = "WSPD", years = 2010,
                                         months = 1:12, bounding_box = bb))

matched <- attach_bathymetry(matched, fetch_bathymetry(bounding_box = bb),
                             c("DEPTH", "SLOPE"))
matched <- attach_climate_index(matched, "NAO")

colSums(is.na(sf::st_drop_geometry(matched)))
```

Each source fails differently at the edges of its record — FVCOM after 2013,
HYCOM after 2015, `LCR` after 2014 — so that last line is worth running before
fitting anything.

## What to cite

Every source here is somebody else's work, and the obligation to cite travels
with the data rather than with this package. **Cite whichever products you
actually fetched from.** Several catalogs carry their references at runtime:

```{r cite, eval = FALSE}
fvcom_archives()$GOM3$reference
hycom_archives()$GLBv53X$reference
ccmp_versions()$`v03.1`$reference
index_dictionary()          # carries the climate index references
```

### Copernicus Marine Service

[Copernicus asks for a specific
form](https://help.marine.copernicus.eu/en/articles/4444611-how-to-cite-copernicus-marine-products-and-services),
including the access date:

> *Product Title*. E.U. Copernicus Marine Service Information (CMEMS). Marine
> Data Store (MDS). DOI: 10.48670/moi-xxxxx (Accessed on DD MMM YYYY)

Reanalysis products, used by default:

| Product | Supplies | DOI |
|---|---|---|
| Global Ocean Physics Reanalysis (GLORYS12V1) | `SST`, `SSS`, `BOTT`, `BOTS`, `UO`, `VO`, `SSH`, `MLD`, `SIC` | [10.48670/moi-00021](https://doi.org/10.48670/moi-00021) |
| Global Ocean Biogeochemistry Hindcast | `CHL_MODEL`, `NPP_MODEL`, `NO3`, `PO4`, `O2`, `PH` | [10.48670/moi-00019](https://doi.org/10.48670/moi-00019) |
| Global Ocean Colour (Copernicus-GlobColour) | satellite `CHL`, `PP`, `DIATO`, `DINO` | [10.48670/moi-00281](https://doi.org/10.48670/moi-00281) |
| Global Ocean Monthly Mean Sea Surface Wind and Stress | `WSPD`, `UWND`, `VWND`, `TAUX`, `TAUY`, `TAU` | [10.48670/moi-00181](https://doi.org/10.48670/moi-00181) |
| Global Ocean Hourly Reprocessed Sea Surface Wind and Stress | the same, with `frequency = "hourly"` | [10.48670/moi-00185](https://doi.org/10.48670/moi-00185) |

`BOTS` is derived from GLORYS's salinity field rather than published by it, so it
carries that product's citation like any other variable taken from it.

The two wind products are separate records with separate DOIs. A study using
monthly wind cites the first; one using hourly wind, or a daily field aggregated
from it, cites the second.

Analysis-and-forecast products, used with `mode = "forecast"`:

| Product | DOI |
|---|---|
| Global Ocean Physics Analysis and Forecast | [10.48670/moi-00016](https://doi.org/10.48670/moi-00016) |
| Global Ocean Biogeochemistry Analysis and Forecast | [10.48670/moi-00015](https://doi.org/10.48670/moi-00015) |

`variable_dataset()` says which product a variable came from, so only the ones
you used need citing. Downloads go through the [Copernicus Marine
Toolbox](https://toolbox-docs.marine.copernicus.eu/), which publishes no DOI of
its own — cite the products.

### Ocean and wind models

- **FVCOM / NECOFS** — Chen C, Beardsley RC, Cowles G (2006). An unstructured
  grid, finite-volume coastal ocean model (FVCOM) system. *Oceanography*
  **19**(1):78–89. <https://doi.org/10.5670/oceanog.2006.92>

- **HYCOM / GOFS 3.1** — Chassignet EP, Hurlburt HE, Smedstad OM, Halliwell GR,
  Hogan PJ, Wallcraft AJ, Baraille R, Bleck R (2007). The HYCOM (HYbrid
  Coordinate Ocean Model) data assimilative system. *Journal of Marine Systems*
  **65**:60–83. <https://doi.org/10.1016/j.jmarsys.2005.09.016>

- **NOAA CEFI / MOM6-COBALT-NWA12** — Ross AC, Stock CA, Adcroft A, Curchitser
  E, Hallberg R, Harrison MJ, Hedstrom K, Zadeh N, Alexander M, Chen W,
  Drenkard EJ, du Pontavice H, Dussin R, Gomez F, John JG, Kang D, Lavoie D,
  Resplandy L, Roobaert A, Saba V, Shin S-I, Siedlecki S, Simkins J (2023). A
  high-resolution physical–biogeochemical model for marine resource
  applications in the northwest Atlantic (MOM6-COBALT-NWA12 v1.0).
  *Geoscientific Model Development* **16**:6943–6985.
  <https://doi.org/10.5194/gmd-16-6943-2023>

  The paper is the model; the data is separately owed an acknowledgement, in
  the form PSL asks for: *Data provided by the NOAA Physical Sciences
  Laboratory, Boulder, Colorado, USA, from https://psl.noaa.gov/cefi_portal/*.
  There is no separate data DOI.

  **Say which release** — `source_of()` records it, as
  `cefi:NWA12-hindcast-r20250715`. Releases extend and revise the record, so
  two results from different releases are not the same numbers. For a forecast,
  say which initialisation and ensemble member as well.

- **CCMP** — Mears C, Lee T, Ricciardulli L, Wang X, Wentz F (2022). Improving
  the Accuracy of the Cross-Calibrated Multi-Platform (CCMP) Ocean Vector Winds.
  *Remote Sensing* **14**(17):4230. <https://doi.org/10.3390/rs14174230>

- **MUR** — Chin TM, Vazquez-Cuervo J, Armstrong EM (2017). A multi-scale
  high-resolution analysis of global sea surface temperature. *Remote Sensing of
  Environment* **200**:154–169. <https://doi.org/10.1016/j.rse.2017.07.029>

- **VIIRS gap-filled chlorophyll** — Liu X, Wang M (2018). Gap filling of
  missing data for VIIRS global ocean color products using the DINEOF method.
  *IEEE Transactions on Geoscience and Remote Sensing* **56**:4464–4476.
  <https://doi.org/10.1109/TGRS.2018.2820423>

- **NASA OB.DAAC** — **two citations are owed, not one.** The mission paper
  below describes the instrument; the *data* has a DOI of its own, under the
  `10.5067` prefix, minted per mission, per suite and per reprocessing — for
  example
  [10.5067/AQUA/MODIS/L3M/CHL/2022](https://doi.org/10.5067/AQUA/MODIS/L3M/CHL/2022)
  for Aqua MODIS chlorophyll.

  The trailing version is not predictable from the suite: for Aqua MODIS the
  `PAR` DOI ending 2022 resolves where the `SST` one ending 2019 does not, and
  a reprocessing changes it. So this
  package points at the DOI rather than shipping a table of them that would go
  stale silently — look yours up at
  <https://www.earthdata.nasa.gov/centers/ob-daac> and cite it alongside the
  mission. `obdaac_sensors()` carries the mission papers at runtime, and
  `source_of()` records which sensor answered, as `obdaac:MODISA-4km`.

  - *SeaWiFS* — McClain CR, Feldman GC, Hooker SB (2004). An overview of the
    SeaWiFS project and strategies for producing a climate research quality
    global ocean bio-optical time series. *Deep-Sea Research II* **51**:5–42.
    <https://doi.org/10.1016/j.dsr2.2003.11.001>
  - *MODIS* — Esaias WE, Abbott MR, Barton I, Brown OB, Campbell JW, Carder KL,
    Clark DK, Evans RH, Hoge FE, Gordon HR, Balch WM, Letelier R, Minnett PJ
    (1998). An overview of MODIS capabilities for ocean science observations.
    *IEEE Transactions on Geoscience and Remote Sensing* **36**:1250–1265.
    <https://doi.org/10.1109/36.701076>
  - *VIIRS* — Wang M, Liu X, Tan L, Jiang L, Son S, Shi W, Rausch K, Voss K
    (2013). Impacts of VIIRS SDR performance on ocean color products. *Journal
    of Geophysical Research: Atmospheres* **118**:10347–10360.
    <https://doi.org/10.1002/jgrd.50793>

Say which archive as well as which model — a value from the HYCOM reanalysis and
one from an operational experiment are not the same run, and `source_of()`
records exactly that.

### Seafloor terrain

- NOAA National Centers for Environmental Information (2022). *ETOPO 2022
  15 Arc-Second Global Relief Model*. <https://doi.org/10.25921/fd45-gt74>
- Pante E, Simon-Bouhet B, Irisson J (2025). *marmap: Import, Plot and Analyze
  Bathymetric and Topographic Data*.
  <https://doi.org/10.32614/CRAN.package.marmap>

`fetch_bathymetry()` requests the 60 arc-second bedrock grid
(`ETOPO_2022_v1_60s_bed`) through `marmap`.

### Climate indices

Two are the published output of specific work and **should be cited when used**:

- **`LCR`** — Jutras M, Dufour CO, Mucci A, Talbot LC (2023). Large-scale control
  of the retroflection of the Labrador Current. *Nature Communications*
  **14**:2623. <https://doi.org/10.1038/s41467-023-38321-y>
- **`AMOC`** — Moat BI, Smeed DA, Rayner D, Johns WE, Smith R, Volkov D, Elipot
  S, Petit T, Kajtar J, Baringer MO, Collins J (2026). *Atlantic meridional
  overturning circulation observed by the RAPID-MOCHA-WBTS array at 26°N from
  2004 to 2024 (v2024.1a)*. British Oceanographic Data Centre, NERC, UK.
  <https://doi.org/10.5285/48d0bf43-0598-ceb2-e063-7086abc062f1>

  BODC mints a new DOI for each release and retires the old one, so this changes
  when RAPID publishes a new version. `index_dictionary()` carries the current
  reference.

The other four are operational products with no single paper behind them. Credit
the provider:

- **`NAO`**, **`AO`** — NOAA Climate Prediction Center.
  <https://www.cpc.ncep.noaa.gov/>
- **`AMO`**, **`PDO`** — NOAA Physical Sciences Laboratory.
  <https://psl.noaa.gov/data/climateindices/>

### Software this is built on

- Pebesma E, Bivand R (2023). *Spatial Data Science: With Applications in R*.
  Chapman and Hall/CRC. <https://doi.org/10.1201/9780429459016> — the `sf`
  reference
- Hijmans R, Brown A, Barbosa M (2026). *terra: Spatial Data Analysis*.
  <https://doi.org/10.32614/CRAN.package.terra>
- Pierce D (2025). *ncdf4: Interface to Unidata netCDF Format Data Files*.
  <https://doi.org/10.32614/CRAN.package.ncdf4> — needed for `AMOC`

`citation("datamatch")` gives this package's own entry, and `citation()` works on
any of the above.

### Keeping these current

Citations go stale without anyone touching them. Data centres reissue a DOI when
a record is superseded and retire the old one, so a reference that was correct
when written stops resolving on its own. The `AMOC` entry here has already been
through that once, when BODC published a newer RAPID release.

Two scheduled workflows watch for it, and open an issue rather than editing
anything, since choosing a replacement is a judgement about which version to
track:

- **Citation check**, quarterly, resolves every DOI cited in the README, the
  catalogs, and `NEWS.md`.
- **Copernicus catalog check**, monthly, compares the variable catalog against
  the live Copernicus catalogue, since dataset identifiers are revised too.

Both run on demand from the Actions tab, and locally:

```{r citation-check, eval = FALSE}
# from the package root
system("Rscript inst/scripts/check_citations.R")
system("Rscript inst/scripts/check_catalog.R")
```

## Satellite or model?

Copernicus serves chlorophyll and primary production from two very different
sources, and both are available:

| Name | Source | Resolution | Trade-off |
|---|---|---|---|
| `CHL`, `PP` | Copernicus-GlobColour, satellite | 4 km | Observed, finer — but surface-only and gappy under persistent cloud |
| `CHL_MODEL`, `NPP_MODEL` | Biogeochemistry reanalysis | 0.25° | Gap-free and depth-resolved — but simulated, and coarser |

The plain names default to satellite, since the values are observed rather than
simulated. Switch to the model versions where cloud gaps would matter more than
resolution.

One caution: **satellite `PP` and model `NPP_MODEL` are not the same quantity.**
`PP` is depth-integrated (mg/m2/day), `NPP_MODEL` volumetric (mg/m3/day).
Substituting one for the other is a units error, not a resolution difference.

Phytoplankton functional types come from the same satellite plankton dataset as
`CHL`, so they can be fetched together:

```{r pft, eval = FALSE}
env <- accessCopernicus(
  vars = c("CHL", "DIATO", "DINO"),   # one request, one dataset
  years = 2003:2017, months = 1:12,
  bounding_box = list(xmin = -76, xmax = -65, ymin = 35, ymax = 45)
)
```

`DIATO` and `DINO` are diatom and dinophyte chlorophyll — the spring-bloom
species large copepods prefer, and the later stratified-water group
respectively.

## Bottom salinity

`BOTS` pairs with `BOTT`, but it is not fetched the same way, because
**GLORYS12V1 does not publish it.** The reanalysis has temperature at the sea
floor and no salinity counterpart. So it is derived: the full salinity column is
fetched and the deepest wet level in each cell kept.

```{r bots-readme, eval = FALSE}
bots <- accessCopernicus(vars = "BOTS", years = 2010:2014, months = 1:12,
                     bounding_box = bb)
#> BOTS is not published by this product. Deriving it from the full 'so' column
#> and keeping the deepest wet level in each cell; the depth used comes back as
#> BOTS_depth.
```

The depth each value came from is returned as `BOTS_depth` rather than left to
be assumed. That matters because the deepest wet *model level* is not the sea
floor: level spacing coarsens with depth, so in deep water the value can sit a
long way above the bottom. In shelf water it is within a few metres.

Two consequences, both deliberate:

- **It must be fetched on its own.** The whole depth column is a different
  request from the single level `SST` wants, so mixing them is an error rather
  than a quiet second download. Call twice and chain `matchData()`.
- **It costs far more.** Roughly fifty levels are downloaded over the same box to
  keep one, so a large box is much slower than the same box of `SST`.

In forecast mode none of this applies. The analysis-and-forecast product
publishes sea-floor salinity outright, so `BOTS` is an ordinary variable there,
fetches alongside `BOTT`, and returns no `BOTS_depth`:

```{r bots-forecast, eval = FALSE}
accessCopernicus(vars = c("BOTT", "BOTS"), years = 2026, months = 8,
             bounding_box = bb, mode = "forecast")
```

The two are the same quantity by construction but not the same number — one is
the deepest level of a 50-level grid, the other Copernicus's own diagnostic — so
a record spanning both modes has a seam in it.

```{r lookup, eval = FALSE}
variable_dictionary("biogeochemical")        # filter by product
variable_dictionary("wind")                  # or just the winds
fvcom_dictionary()                           # the FVCOM catalog, for accessFVCOM()
fvcom_archives()                             # which FVCOM archives are built in
hycom_dictionary()                           # the HYCOM catalog, for accessHYCOM()
hycom_archives()                             # which HYCOM archives can be read
hycom_covering("2019-06-15")                 # which of them span a given date
ccmp_dictionary()                            # the CCMP catalog, for accessCCMP()
ccmp_versions()                              # which CCMP versions can be read
erddap_dictionary()                          # MUR and VIIRS, for accessERDDAP()
erddap_datasets()                            # which ERDDAP datasets ship
variable_dataset(c("SST", "CHL"))            # which dataset each comes from
as.data.frame(variable_dictionary())$description  # full descriptions
```

## Error messages

Most of what this package refuses to do, it refuses deliberately. These are the
messages you are most likely to meet, and what each one means.

**`Could not find the Copernicus Marine client`**

R cannot see `copernicusmarine` on its `PATH`. Common when it lives in a conda
environment that RStudio does not inherit. Point at it directly:

```{r ts-client, eval = FALSE}
options(datamatch.copernicusmarine = "~/miniconda3/bin/copernicusmarine")
```

**`These variables come from different Copernicus datasets`**

Expected, not a fault. `SST` is physics and `CHL` is biogeochemistry, on
different grids. Fetch them separately and chain `matchData()`, as in
[Putting several together]. `variable_dataset()` shows which dataset each
variable comes from.

In forecast mode this happens more often, because the forecast products split
variables across more datasets than the reanalysis does. `SST` and `UO` share a
dataset as reanalysis but not as forecast.

**`Copernicus publishes no daily dataset for: UWND, VWND`**

Expected. This wind is published hourly or monthly and nothing between. Fetch
`frequency = "hourly"` and aggregate with `upscale_time(to = "day")`. The same
message for `PH`, `PP`, `DIATO` or `DINO` means the opposite — those are monthly
composites, so fetch them monthly.

**`Copernicus publishes no hourly dataset for: WSPD`**

`WSPD` and `TAU` are magnitudes the hourly wind product does not carry. Fetch
`UWND` and `VWND` hourly and compute the magnitude, or take these monthly.

**`BOTS must be fetched on its own`**

Bottom salinity is derived from the whole depth column, which is a different
request from the single level a surface variable wants. Call `accessCopernicus()`
once for `BOTS` and once for the rest, then chain `matchData()`. See
[Bottom salinity].

**`These variables sit on different parts of the FVCOM mesh`**

Scalars sit on mesh nodes and velocities on element centroids, so the two kinds
cannot be read together. Call `accessFVCOM()` once for each and chain
`matchData()`.

**`The CEFI hindcast saves no daily SST`**

Expected. CEFI's daily output is biogeochemistry only — `CHL`, `NO3`, `PH`,
`PCO2`, `PHYC`, `MESOZOO` and `BOTO2`. Everything else is monthly, and the
message lists what is available daily.

**`The bounding box selects no CEFI cells`**

CEFI is a regional model, not a global one. The Northwest Atlantic domain runs
roughly 98°W–36°W and 5°N–58°N, and the message gives the actual extent.
Longitudes are negative west, and the regridded files are already on that
convention, so no conversion is needed in either direction.

**`CEFI's 'seasonal_forecast' cannot be read through this package`**

Not a fault in your call. Those files carry a 64-bit integer coordinate that
DAP2 has no type for, so the server returns HTTP 500 to every OPeNDAP request;
the DAP4 endpoint describes them and then reads them wrongly, crashing R. The
files download whole from the THREDDS `fileServer` endpoint and open correctly
once local, at roughly 260 MB per variable per initialisation.

**`init` / `member` `is required for CEFI's decadal forecast`**

Deliberate. A decadal file is ten years from one January and there are sixty of
them, so which initialisation you mean is a question about your analysis. Each
holds ten ensemble members, and averaging them is a modelling decision this
package will not make silently — read the members you want and combine them
yourself, so the combination is visible in your code.

**`Reading OB.DAAC needs an Earthdata Login`**

The one account this package cannot work around. Register at
<https://urs.earthdata.nasa.gov/users/new>, generate an appkey at
<https://oceandata.sci.gsfc.nasa.gov/appkey/>, and put it in `~/.Renviron` as
`EARTHDATA_APPKEY=`. A `~/.netrc` entry for `urs.earthdata.nasa.gov` works too.

**`the server returned a login page rather than the file`**

The credential is missing, wrong, or revoked. NASA answers an unauthenticated
request with HTTP 200 and the login page rather than a 401, so this is caught by
checking the bytes; nothing is written to disk. A date the sensor did not return
lands here too, since the two are indistinguishable from the response.

**`MODISA does not publish: NFLH`** *(or similar)*

The sensor does not carry that suite. SeaWiFS measures no temperature at all;
only MODIS carries `NFLH`; VIIRSJ2 has no SST yet. The message lists what that
sensor does carry, and `obdaac_sensors()` has the full table.

**`The download did not return: uo, vo`**

Those variables were requested but are not in the returned file. Either the
dataset does not serve them, or they are unavailable at the requested depth or
date. The message lists what did arrive.

**`The depth range returned several model levels`**

`depth` spanned more than one model level, so a variable came back on several
layers and there is no way to know which was wanted. Request a single level with
`depth = c(0, 1)`.

**`Copernicus download failed after 3 attempt(s)`**

The request was retried and kept failing. The message carries the client's own
output. Rapid repeated downloads can also be rate-limited, in which case waiting
and re-running works.

**`Expected N variable column(s) but the download returned M`**

From a version before the layer-matching fix. Update the package: this message
blamed the depth range for causes it could not distinguish, and the two real ones
now report themselves. See `NEWS.md`.

**Values in the wrong range for their column name**

Salinity around 32 in an `SST` column means you are on a version predating the
layer-ordering fix, where multi-variable downloads could be mislabelled silently.
Update and re-fetch. `NEWS.md` describes what was affected and how to check.
