---
title: "Working with what comes back"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with what comes back}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(datamatch)
```

Every access function returns the same thing: an `sf` point object with one row
per cell per time step. This is what to do with it — moving between grids and
time steps, filling the holes satellite products leave, and looking at what
arrived before trusting it.

None of it is source-specific. A grid from Copernicus, FVCOM, HYCOM, CCMP or
ERDDAP resamples and plots the same way.

## Resampling

Four functions, one pair per axis, plus one for gaps. Each takes a `method`,
because there is no single right way to change resolution:

| Function | What it does |
|---|---|
| `upscale_grid()` | Aggregates onto a coarser grid — `mean`, `median`, `min`, `max`, `sum`, `mode` |
| `downscale_grid()` | Interpolates onto a finer grid — `nearest`, `bilinear`, `cubic`, `idw` |
| `upscale_time()` | Aggregates onto a coarser time step — hourly to daily, daily to monthly, monthly to annual |
| `downscale_time()` | Interpolates onto a finer time step — `step`, `linear`, `spline` |
| `fill_satellite_gaps()` | Substitutes another covariate wherever the first is missing |

```{r resample, eval = FALSE}
chl <- accessCopernicus(vars = "CHL", years = 2010, months = 1:12, bounding_box = bb)
sst <- accessCopernicus(vars = "SST", years = 2010, months = 1:12, bounding_box = bb)

# Satellite CHL (4 km) onto the physics grid, so the two can be modelled together
chl_on_sst <- upscale_grid(chl, to = sst)

# Or onto a stated resolution
chl_quarter <- upscale_grid(chl, to = 0.25, method = "median")

# Daily data to monthly means; monthly back out to daily
monthly <- upscale_time(daily_sst, to = "month")
daily   <- downscale_time(monthly, to = "day")
```

The target is either a resolution or **another object whose grid to adopt**. The
second form is what puts two products on one grid.

A resolution is anchored to the origin, not to your data's corner. So two
products upscaled to 0.25° land on the *same* cells, instead of half a cell
apart.

**The right treatment differs by covariate**, which is why this is per-covariate
rather than one setting for everything. Ocean-colour chlorophyll is a 4 km optical
retrieval riddled with cloud gaps. Physics is a 0.083° model with no gaps at all.

Averaging chlorophyll up to the physics grid summarises values that were really
measured. Interpolating physics down to 4 km invents structure. Those are not the
same operation, and no single choice can be right for both.

### Aggregating loses detail, which is the safe direction

`mean` is the default. The others exist because "the value of this coarse cell"
is not one question:

- `median` resists the retrieval outliers at satellite cloud edges.
- `min` of depth is the shallowest point a cell contains, not its average.
- `sum` is for per-cell totals. It is meaningless for a concentration.
- `sd` over a period is variability, which is a covariate in its own right.

Pass one method for everything, or a named vector per variable:

```{r methods, eval = FALSE}
upscale_grid(env, to = 0.25, method = c(CHL = "median", DEPTH = "min"))
```

**Partly-covered periods and cells come back `NA` by default.** A mean over the
four days of January you happened to fetch is not a January mean, and nothing in
the number itself says so.

`min_coverage` defaults to 0.5. It measures against the steps a period *has* (31
for January), not the steps you downloaded. Set `min_coverage = 0` to aggregate
whatever is present. Set `keep_counts = TRUE` to see what was behind each value.

### Interpolating adds cells, not information

This is the direction to be careful in. A 0.25° field rendered at 4 km has 4 km
*cells*, but it still resolves nothing below 0.25°. None of these methods consult
any other data source, so there is nowhere for real fine-scale structure to come
from.

That is why the defaults are the blunt ones. `nearest` and `step` replicate the
source value, leaving the result visibly blocky at the source resolution, so the
output looks like what it is. `bilinear`, `cubic`, `linear`, and `spline` return
smooth fields that look like finely-resolved measurements and are not.

There is a further trap specific to the time axis:

> **`linear` and `spline` do not preserve the period mean.** Interpolate twelve
> monthly means to daily values, average those days back up, and you will not
> recover the months you started from — in the package's own tests a March mean
> of 15.0 comes back as 14.83. Any budget or total computed from such a series
> inherits that error. `step` preserves it exactly.

`idw` is the exception worth reaching for. Alone among the spatial methods, it
fills across holes rather than propagating them. That makes it useful on gappy
satellite data, where `fill_satellite_gaps()` is not an option.

Resampled output keeps the `YEAR`/`MONTH`/`DAY` stamping convention, so
`matchData()` reads its resolution back correctly and can be used on the result
directly.

## Filling satellite gaps

```{r gapfill, eval = FALSE}
chl_sat <- accessCopernicus(vars = "CHL", years = 2010, months = 1:12, bounding_box = bb)
chl_mod <- accessCopernicus(vars = "CHL_MODEL", years = 2010, months = 1:12, bounding_box = bb)

filled <- fill_satellite_gaps(chl_sat, chl_mod, c(CHL = "CHL_MODEL"))
table(filled$CHL_source)
#> satellite     model   missing
#>      8214      1902         0
```

The two sources measure the same quantity in different ways. They are not
interchangeable. So the seam is left visible rather than smoothed over. Nothing
is rescaled by default, and a `<var>_source` column records where each value came
from.

`rescale = TRUE` matches the model's mean and variance to the satellite's over
the cells where both exist. That reduces the step, at the cost of altering the
model values.

## Looking at the data

Four plots, in the order they are usually wanted. All use base graphics, so they
need no extra packages. Each returns the data it drew, so the numbers behind a
picture are always available.

```{r plot-env, eval = FALSE}
env <- accessCopernicus(vars = c("SST", "CHL"), years = 2010, months = 1:12,
                    bounding_box = bb)

plot_env(env)                                # first variable, first time step
plot_env(env, "CHL", time = c(MONTH = 6))    # June chlorophyll
```

`plot_env()` maps one variable for one time step. It is the first thing worth
doing after a download. It is also the fastest way to catch a bounding box that
landed somewhere unintended, a variable that is entirely `NA`, or a depth range
that returned the wrong level.

Data are drawn as a raster, so gaps read as holes rather than absent dots. The
time step is named in the title, so a map is never ambiguous about which month it
shows.

```{r plot-coverage, eval = FALSE}
plot_coverage(env)
```

`plot_coverage()` is the one to run before trusting a satellite series. It plots
the fraction of cells carrying a value in each time step. On ocean colour the
shape is stark: near-complete in summer, a quarter of the grid in winter.

That picture decides three things. Whether a monthly mean is worth having.
Whether `fill_satellite_gaps()` is worth the seam it introduces. And where to set
`min_coverage` when aggregating.

```{r plot-series, eval = FALSE}
plot_series(env)                     # one panel per variable
plot_series(env, "SST", fun = max)   # the warmest cell each month
```

`plot_series()` reduces each time step to one number over the study area. That is
how a seasonal cycle, a trend, or a step change at a product boundary becomes
visible.

The interquartile range across cells is shaded behind the line. A mean alone
hides the difference between a uniformly warm month and one that is warm inshore
and cold offshore.

```{r plot-matched, eval = FALSE}
matched <- matchData(observations, env)

plot_matched(matched[matched$MONTH == 7, ], "SST")
```

`plot_matched()` shows what the join actually produced. Observations that matched
nothing are drawn as open circles rather than dropped. A cluster of them is
usually the real finding: observations outside the environmental data's extent,
or in a period it does not cover.

Subset to one period first, as above. The colour scale spans everything passed
in. Plot a whole year of a seasonal variable and the seasons mix together, so the
map reads as noise. A warm February inshore point and a cool August offshore one
can take the same colour.

## Downloads run in parallel

**`n_workers` controls this, and it defaults to 4** — parallel downloading is on
without being asked for. A Copernicus subset request spends nearly all of its
time waiting on the API rather than on your machine, so several requests in
flight take barely longer than one.

Two months of daily SST and SSS over a one-degree box — 59 downloads — each run
cold-cache, all three back to back in one session:

| `n_workers` | Elapsed | |
|---|---|---|
| 1 | 295 s | |
| 4 (the default) | 80 s | **3.7× faster** |
| 6 | 51 s | **5.8× faster** |

All three returned identical data. The gain scales with workers rather than with
cores, because the wait is the network rather than your machine. How much you see
depends on the API's mood as much as anything, so treat these as the shape of it
rather than a promise.

```{r quickstart-workers, eval = FALSE}
# The default. 15 years x 12 months is 180 downloads, four at a time
accessCopernicus(vars = "SST", years = 2003:2017, months = 1:12, bounding_box = bb)

accessCopernicus(vars = "SST", years = 2003:2017, months = 1:12, bounding_box = bb,
             n_workers = 8)   # more at once
accessCopernicus(vars = "SST", years = 2003:2017, months = 1:12, bounding_box = bb,
             n_workers = 1)   # one at a time
```

**This is not a daily-only feature.** Any fetch spanning more than one time step
is more than one download, so a long monthly record benefits as much as a daily
one — the 180 downloads above are monthly.

Two things it deliberately does not do. It does not re-download what you already
have: files in the cache are read straight from disk, and a fully cached call
starts no workers at all. And it does not parallelise across calls — each
`accessCopernicus()` call is one dataset, so fetching SST and CHL is two calls, run
one after the other with each parallel inside itself.

The limit is the service, not your cores, so raising `n_workers` far past 8
mostly earns rate limiting. A day that fails does not abandon the others: every
day is attempted, the ones that succeeded stay cached, and the error names each
day that failed — so re-running the same call retries only those.

`variable_dataset(c("SST", "CHL"))` tells you which product each variable comes
from, so you can plan the calls before making them. See [Putting it
together](#putting-it-together) for the same pattern with seafloor and
basin-scale covariates added.
